A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-10-25 thesis/thesis.tex
apple franck comments

\begin{document}
\title{A Wait-free Queue with Polylogarithmic Step Complexity}
\author{Hossein Naderibeni \\supervised by Eric Ruppert}
\maketitle

\begin{abstract}
In this work, we introduce a novel linearizable wait-free queue implementation.
Linearizability and lock-freedom are standard requirements for designing shared data structures. To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case called the CAS Retry Problem. We show our algorithm avoids this problem and has worst case running time better than prior lock-free queues. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm, is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
\end{abstract}

\tableofcontents
\pagebreak
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure to work with. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can make progress at the time data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored. \texttt{Dequeue} removes and returns the first element in the queue. If the queue is empty it returns \texttt{null}. In a concurrent version of queue data structure, operations do not happen one at a time. The question that may arise is ``What properties matter for implementation of a shared data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes. Our model is an asynchronous shared-memory distributed system of $p$ processes. Herlihy showed \cite{10.1145/114005.102808} that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy on how powerful objects are to solve the consensus problem among processes. Objects like LL/SC, CAS can be used to reach consensus among any number of processors. A CAS object provides an atomic Compare \& Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}, otherwise it returns \nf{false}.  In this work we use Compare \& Swap (\nf{CAS}) objects to synchronize among  processes.

\bf{Concurrent execution: I didn't understed interleaving steps of processes}

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable. \texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can be occurred before termination of \texttt{Enqueue(A)} and return \nf{null} or can be occurred after termination of \texttt{Enqueue(A)} and return \texttt{A}.}
\end{figure}

@@

We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps. We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized. Processes can do queries to get information about the linearization stored in the root. To avoid the CAS Retry problem, when a process is propagating its operation to a node we create a helping mechanism that ensures operations are propagated after at most 2 \nf{CAS} invocations. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized. \nf{CAS} instructions cost more than other instructions for processor. Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case. We use unbounded memory space in our queue, we present a way to reduce the memory size to the total number of operations. We do not address garbage collection but we describe a way we think it can be handled.

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of related word done in the area and the motivation of our implementation. Previous lock-free queues and their common problem is presented in Section 2.1. In Section 2.2 we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 is summoned with a lower bound on the amortized time complexity of shared queues.

@@

Chapter 5 is devoted to performance analysis for our implementation. We analyze the number of \nf{CAS} instructions out algorithm invokes and the worst-case and amortized time running of the queue. In the end we prove our queue is wait-free.

@@

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. The queue is stored in a shared array where head and tail pointers point to the current elements in the queue. When the array is occupied an empty array is linked to the array and tail pointers are updated. A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array. Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}). A process updates the position of the pointers after it does an operation. One process might go to sleep before setting the pointers, so the pointers might be behind their real places. They mention how to scan the arrays to update pointers while doing an operation. A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the step (and \nf{CAS}) complexity remains $\Omega(p)$. Their design is lock-free but not wait-free.

@@

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology. Their work is based on a circular queue using bounded memory. When a process wishes to do an enqueue or a dequeue, it starts two paths. One is faster but its running time is not bounded the other one is slower but has a time bound. They show that these two paths do not effect each other and the queue remains consistent. If a process has no progress other processes help its slow path to finish. The helping phase suffers from CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help. Because of this, the worst-case complexity cannot be better than $\Omega(p)$.

@@

David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}. Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinte memory.
The author states that to reduce memory usage to be bounded the time per operation increases to linear.

Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}. We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a a poly-logarithmic multi-enqueuer multi-dequeuer queue. Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.

@@

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Incs added by processes. When a process wants to do Fetch\&Inc it adds a Fetch\&Inc to the tree and return the number of elements in the tree. There are  some similarities between designing a queue and a Fetch\&Inc object. A Fetch\&Inc object can be constructed from a queue. The algorithm by Ellen and Woelfel is interesting because it is the first wait-free data structure achieving poly-log implementation in a while.

\subsection{Attiya--Fouren Lower Bound}
Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in time complexity of concurrent queues. Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard ti $c$, the number of concurrent processes. Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is that their result does not contradict ours and we manage to reach poly-log time complexity.

@@

It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
Instead we use an implicit representation of the operations propagated together. Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}). Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}s. A block stores information about the operations propagated by a \nf{Refresh}. It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure. See Figure \ref{fig:block} for an example. A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in children into a new block and puts it in the parent's blocks. We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently. Thus, there are at most $p$ operations in a block. Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.

@@

So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on linearization ordering we can design a universal construction; for given object $O$ we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$. However, this approach is not enough for an efficient queue.
We show that we can build an efficien queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization. We explain how to implement (1) and (2) in poly-logarithmic steps.

After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblocks of $B$ in the descendent of the root that contain that $i$th operation. When we reach a block in a leaf, the operation is explicitly stored there. To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block. We also store pointers to determine the range of subblocks of a block to make the binary search faster. In each block, we store the prefix sum of operations from the left child and from the right child. Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figure \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.

@@

In an execution on a queue where no dequeue operation returns\nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue. In the general case a dequeue returns\nf{null} if and only if the size of the queue after the previous operation is 0. We refer to such a dequeue as a \it{null dequeue}. If the the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue. Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues up to block B. So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.

@@

\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively.
Now, we claimed that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}. We apply this approach to blocks; if we store the size of the queue after each block of operations happens, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.

@@

\subsection{Details of the Implementation}
Pseudocode for the queue implementation is given in Section \ref{algQ}. It uses the following two types of objects.
\paragraph{\tt{Node}} 
 In each \texttt{Node} we store pointers to its parent and children, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.

\paragraph{\tt{Block}}
 The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf. If it is in a leaf, we use a \nf{LeafBlock} which simply stores one operation. If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$. The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the block prior to $B$ ended. In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}. The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child. The other fields' semantics are similar. The \nf{end\sub{left}} and \nf{end\sub{right}} field store the last subblock of a block in the left and the right child, respectively. The approximate index of the superblock of non-root blocks is stored in their \nf{super} field. The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.

@@

\paragraph{\tt{FindResponse($b, i$)}}
To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not. If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}. If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is response to the \nf{Dequeue}. Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block) we find the block and position containing the \nf{Enqueue} operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.

@@

\paragraph{\tt{Propagate()}}
\nf{Propagate} on node $n$ uses the double refresh idea described in Section \ref{DescriptQ} and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}. Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.

\paragraph{\tt{Refresh()}}
The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}. The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}. The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$. Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children. \nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recent appended blocks. The reason behind this helping is explained later when we discuss \nf{IndexDequeue}. After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}. Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate and it is unnecessary to add an empty block to the tree. Later we will use the fact that all blocks contain at least one operation. Line \ref{cas} tries to install \nf{new}. If it was successful all is good. If not, it means someone else has already put a block in $n$\nf{.blocks[h]}. In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.

\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
The block \nf{new} is created in Line 333 and its fields are filled similarly for both left and right directions. The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}. Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child. Then \nf{sum\sub{enq-dir}} is computed from sum of the the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}. The field \nf{sum\sub{deq-dir}} is computed similarly. Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.

@@

A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block in the root of the $i$th \nf{Dequeue} of the $b$th block of $n$. Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}. Let $par$ be $n$\nf{.parent}. Without loss of generality assume for the rest of this section a ${n}$ is the left child of $par$. Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ in its created block in Line \ref{invokeCreateBlock}. Let $j$ be the index of the block that $R_{par}$ put in $par$\nf{.blocks}.

>>>>

\begin{document}
\title{A Wait-free Queue with Poly-logarithmic Step Complexity}
\author{Hossein Naderibeni \\supervised by Eric Ruppert}
\maketitle

\begin{abstract}
In this work, we introduce a novel linearizable wait-free queue implementation.
Linearizability and lock-freedom are standard requirements for designing shared data structures. To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case called the CAS Retry Problem. We show that our algorithm avoids this problem and has worst case running time better than prior lock-free queues. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm, is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
\end{abstract}

\tableofcontents
\pagebreak
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure to work with. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can make progress at the time the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored. \texttt{Dequeue} removes and returns the first element in the queue. If the queue is empty it returns \texttt{null}. In a concurrent version of the queue data structure, operations do not happen one at a time. The question that may arise is ``What properties matter for implementation of a shared data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes. Our model is an asynchronous shared-memory distributed system of $p$ processes. Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy on how powerful objects are to solve the consensus problem among processes. Objects like LL/SC, CAS can be used to reach consensus among any number of processors. A CAS object provides an atomic Compare \& Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}, otherwise it returns \nf{false}.  In this work we use Compare \& Swap (\nf{CAS}) objects to synchronize among  processes.

% \bf{Concurrent execution: I didn't understed interleaving steps of processes}

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable. \texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occurr before termination of \texttt{Enqueue(A)} and return \nf{null} or can occurr after termination of \texttt{Enqueue(A)} and return \texttt{A}.}
\end{figure}

@@

We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps. We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized. Processes can do queries to get information about the linearization stored in the root. We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated after at most 2 \nf{CAS} invocations. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized. \nf{CAS} instructions cost more than other instructions for processor. Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case. We use unbounded memory space in our queue, we present a way to reduce the memory size to the total number of operations. We do not address garbage collection but we describe a way we think it can be handled.

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of related work done in the area and the motivation of our implementation. Previous lock-free queues and their common problem is presented in Section 2.1. In Section 2.2 we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.

@@

Chapter 5 is devoted to performance analysis for our implementation. We analyze the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue. In the end we prove our queue is wait-free.

@@

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. The queue is stored in a shared array where head and tail pointers point to the current elements in the queue. When the array is full, an empty array is linked to the array and tail pointers are updated. A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array. Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}). A process updates the position of the pointers after it does an operation. One process might go to sleep before setting the pointers, so the pointers might be behind their real places. They mention how to scan the arrays to update pointers while doing an operation. A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the step (and \nf{CAS}) complexity remains $\Omega(p)$. Their design is lock-free but not wait-free.

@@

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology. Their work is based on a circular queue using bounded memory. When a process wishes to do an enqueue or a dequeue, it starts two paths. One is faster but its running time is not bounded and the other one is slower but has a time bound. They show that these two paths do not effect each other and the queue remains consistent. If a process makes no progress other processes help its slow path to finish. The helping phase suffers from CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help. Because of this, the worst-case complexity cannot be better than $\Omega(p)$.

@@

David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}. Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinite memory.
The author states that to reduce memory usage to be bounded the time per operation increases to linear.

Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}. We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a poly-logarithmic multi-enqueuer multi-dequeuer queue. Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.

@@

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes. When a process wants to do Fetch\&Inc it adds a Fetch\&Inc to the tree and returns the number of elements in the tree. There are some similarities between designing a queue and a Fetch\&Inc object. A Fetch\&Inc object can be constructed from a queue. The algorithm by Ellen and Woelfel is interesting because it is the first wait-free data structure achieving poly-log implementation after~\cite{DBLP:conf/podc/ChandraJT98}.

\subsection{Attiya--Fouren Lower Bound}
Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues. Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes. Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is that their result does not contradict ours and we manage to reach poly-log time complexity.

@@

It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
Instead we use an implicit representation of the operations propagated together. Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}). Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instruction on fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}s. A block stores information about the operations propagated by a \nf{Refresh}. It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure. See Figure~\ref{fig:block} for an example. A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in children into a new block and puts it in the parent's blocks. We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently. Thus, there are at most $p$ operations in a block. Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.

@@

So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on linearization ordering we can design a universal construction; for a given object we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$. However, this approach is not enough for an efficient queue.
We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization. We explain how to implement (1) and (2) in poly-logarithmic steps.

After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblocks of $B$ in the descendent of the root that contain that $i$th operation. When we reach a block in a leaf, the operation is explicitly stored there. To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block. We also store pointers to determine the range of subblocks of a block to make the binary search faster. In each block, we store the prefix sum of operations from the left child and from the right child. Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figure \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.

@@

In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue. In the general case a dequeue returns \nf{null} if and only if the size of the queue after the previous operation is 0. We refer to such a dequeue as a \it{null dequeue}. If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue. Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation. So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.

@@

\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
Now, we claimed that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}. We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.

@@

\subsection{Details of the Implementation}
Pseudocode for the queue implementation is given in Section \ref{algQ}. It uses the following two types of objects.
\paragraph{\tt{Node}} 
 In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.

\paragraph{\tt{Block}}
 The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf. If it is in a leaf, we use a \nf{LeafBlock} which simply stores one operation. If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$. The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended. The right subblocks of $B$ are also defined similar. In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}. The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child. The other fields' semantics are similar. The \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of the last subblock of a block in the left and the right child, respectively. The approximate index of the superblock of non-root blocks is stored in their \nf{super} field. The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.

@@

\paragraph{\tt{FindResponse($b, i$)}}
To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not. If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}. If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}. Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block) we find the block and position containing the \nf{Enqueue} operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.

@@

\paragraph{\tt{Propagate()}}
\nf{Propagate} on node $n$ uses the double refresh idea described earlier and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}. Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.

\paragraph{\tt{Refresh()} and \tt{Advance()}}
The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}. The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}. The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$. Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children. \nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recent appended blocks. The reason behind this helping is explained later when we discuss \nf{IndexDequeue}. After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}. Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate and it is unnecessary to add an empty block to the tree. Later we will use the fact that all blocks contain at least one operation. Line \ref{cas} tries to install \nf{new}. If it was successful all is good. If not, it means someone else has already put a block in $n$\nf{.blocks[h]}. In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.

\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
The block \nf{new} is created in Line \ref{initNewBlock} and its fields are filled similarly for both left and right directions. The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}. Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child. Then \nf{sum\sub{enq-dir}} is computed from the sum of the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}. The field \nf{sum\sub{deq-dir}} is computed similarly. Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.

@@

A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block in the root of the $i$th \nf{Dequeue} of the $b$th block of $n$. Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}. Let $par$ be $n$\nf{.parent}. Without loss of generality assume for the rest of this section that ${n}$ is the left child of $par$. Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ in its created block in Line \ref{invokeCreateBlock}. Let $j$ be the index of the block that $R_{par}$ puts in $par$\nf{.blocks}.
